---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# CVR_Harvard-MIT

<!-- badges: start -->
<!-- badges: end -->

The goal of CVR_Harvard-MIT is to coordinate the data combination of the "Harvard" and "MIT" team's CVR data. 


The goal of `01_compare-pres-counts` is to count presidential candidates votes between the two teams. `share/viz.html` creates a dashboard to compare.


The goal of `02_tab-medsl` is for Shiro to tabulate MEDSL's export and check it with the Harvard team's counts. 

# Version of Main Dataset

The large CVR datasets are **not** tracked in git. They are instead stored in each analyst's local directories. 

The **latest version** of MEDSL's full CVR data is from @mreece13:

```{r}
fs::dir_info("data/MEDSL/cvrs_statewide") |> 
  dplyr::select(path, birth_time, modification_time)
```

The MD5 hash of the underlying zipfile is

```{bash}
md5 data/MEDSL/cvrs_statewide.zip
```




# Meeting Notes


Notes from 2024-01-05 meeting: 

- aim to end up with columns district id, contest, candidate (name + party). NOT precinct. 
- can withhold counties where counts are off by X% or where there's only one vote method
- more challenging: trace back ID row to raw format. full R only script for replication 
